My wonderful paper

### 2024-12-05 _manuscript/_tex/index.tex
update the linear regression example

The rest of the paper is organized as follows:
Section~\ref{sec-lit-review} reviews the concept of diagnosing
unexpected outcomes and data quality checks. Section~\ref{sec-plan}
introduces the concept of analysis plan, illustrated with a toy example
on fitness step count. Section~\ref{sec-method} describes the procedure
for selecting optimal check combinations to predict unexpected outcomes.
Section~\ref{sec-examples} applies this selection procedure to a linear
regression example examining the effect of PM10 on mortality.
Section~\ref{sec-discussion} discusses a few key considerations and
Section~\ref{sec-conclusion} concludes the paper.

@@

A regression example is produced to illustrate the test selection
process for analyzing the effect of PM10 on mortality. The example
demonstrates how the test process can be used to select cutoff value in
the unit tests and how the test can be used to diagnose the unexpected
results.

\subsection{Linear regression}\label{sec-linear-reg}

Consider a linear regression model to study the effect of PM10 on
mortality (provide context of using PM10 to study mortality). From the
literature, analysts may expect a significant (p-value \(\le\) 0.05)
PM10 coefficient in the linear model after considering the temperature
confounding {[}TODO add reference{]}. This expectation can be framed
into a unit test: the test fails, labelled as 1, if the p-value of the
PM10 coefficient is greater than 0.05. And we have devised three unit
tests to consider the 1) sample size, 2) strength of correlation between
mortality and PM10, and 3) strength of correlation between mortality and
temperature. {[}sample size matters because \ldots, correlation between
mortality and PM10 matters because\ldots{]}. However, it is not clear
how to set the cutoff value for these tests to determine a failure. Here
a list of tests are created in Table~\ref{tbl-tests}.

\begin{longtable}[]{@{}l@{}}
\caption{A list of unit tests created for the linear regression model of
mortality on PM10 and temperature. The tests are based on the sample
size, correlation between mortality and PM10, and correlation between
mortality and temperature. The cutoff values for each test are specified
to determine a failure.}\label{tbl-tests}\tabularnewline
\toprule\noalign{}
the test fails if \ldots{} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
the test fails if \ldots{} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
sample size less than or equal to 200 \\
sample size less than or equal to 400 \\
sample size less than or equal to 600 \\
sample size less than or equal to 800 \\
mortality-PM10 correlation less than -0.03 \\
mortality-PM10 correlation less than -0.04 \\
mortality-PM10 correlation less than -0.05 \\
mortality-PM10 correlation less than -0.06 \\
mortality-temperature correlation greater than -0.3 \\
mortality-temperature correlation greater than -0.35 \\
mortality-temperature correlation greater than -0.4 \\
mortality-temperature correlation greater than -0.45 \\
\end{longtable}

To generate replicate of the data, we first simulate the correlation
matrix for the three variables: mortality, PM10, and temperature with
the following vector:

\begin{itemize}
\tightlist
\item
  mortality and PM10: -0.03 to -0.1
\item
  mortality and temperature: -0.2 to -0.4
\item
  PM10 and temperature: 0.2 to 0.4
\end{itemize}

We then use a Gaussian copula to generate a multivariate normal
distribution based on the specified correlation matrix and sample size.
The multivariate normal distribution is transformed using the normal
CDF, followed by the inverse CDF of the assumed distributions of the
three variables. Various distributions are fitted to the variables,
including poisson and negative binomial for mortality; gamma,
log-normal, exponential, weibull, and normal for pm10 and temperature;
and beta for pm10 after rescaling the data to \([0-1]\). AIC is used to
determine the best distribution fit for each variable with qq plot
presented in Figure~\ref{fig-dist-fit} to evaluate the fit. The
following distributions and parameters, estimated from the original
data, are used to simulate the three variables:

\begin{itemize}
\tightlist
\item
  mortality, negative binomial distribution with size = 74 and mu = 183
\item
  PM10, beta distribution with shape1 = 4.21 and shape2 = 11.67, then
  multiply by 100
\item
  temperature, weibull distribution with shape = 3.8 and scale = 61
\end{itemize}

@@

\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedleft
tests
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
precision
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
recall
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
overlapping
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
independence
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
harmonic
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
arithmetic
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
1 & 0.713 & 0.662 & 0 & 1 & 0.767 & 0.792 \\
2 & 0.906 & 0.748 & 0 & 1 & 0.872 & 0.885 \\
3 & 0.789 & 0.733 & 0 & 1 & 0.826 & 0.841 \\
4 & 0.908 & 0.794 & 0 & 1 & 0.893 & 0.901 \\

@@

A total of 12 tests with four cutoff values each for the three types of
tests are used to fit the logic regression and calculate the precision
and recall score for predictability of an unexpected outcome (p value
\textless{} 0.05). Figure~\ref{fig-linear-reg-tree} shows the logic
regression tree from the fitted model and Table~\ref{tbl-linear-reg}
prints the numerical summary of three selected single test and their
combined test found by the logic regression. The logic regression model
is capable of picking up the best cutoff value for each type of tests
and yields the following three tests:

\begin{itemize}
\tightlist
\item
  sample size less than or equal to 600
\item
  mortality-temperature correlation greater than -0.35
\item
  mortality-PM10 correlation less than -0.05
\end{itemize}

The combined test rule, (mortality-tempearture correlation AND sample
size) OR mortality-PM10 correlation, has the highest precision, recall,
and independence, suggesting that it is the most effective test for
diagnosing unexpected results in the linear regression model given our
assumption on the data. The tree structure suggests obtaining the
mortality-PM10 correlation is as important as obtaining a combined
relationship between mortality-temperature correlation and sample size
in diagnosing unexpected results in the linear regression model. And
based on our assumption of the variable distribution from the data
simulation, we can obtain a precision of 0.908 and a recall of 0.794 for
the unexpected outcome. These provides check points for the analysts to
diagnose the unexpected results and to improve the model.

>>>>

The rest of the paper is organized as follows:
Section~\ref{sec-lit-review} reviews the concept of diagnosing
unexpected outcomes and data quality checks. Section~\ref{sec-plan}
introduces the concept of analysis plan, illustrated with a toy example
on fitness step count. Section~\ref{sec-method} describes the procedure
for selecting optimal check combinations to predict unexpected outcomes.
Section~\ref{sec-pm10-mortality} applies this selection procedure to a
linear regression example examining the effect of PM10 on mortality.
Section~\ref{sec-discussion} discusses a few key considerations and
Section~\ref{sec-conclusion} concludes the paper.

@@

A regression example is produced to illustrate the test selection
process for analyzing the effect of PM10 on mortality. The example
demonstrates how the test process can be used to select cutoff values in
the unit tests and how the test can be used to diagnose an unexpected
PM10 coefficient from the generalized linear model.

\subsection{Effect of PM10 on mortality}\label{sec-pm10-mortality}

Consider a generalized linear model (GLM) to study the effect of PM10 on
mortality {[}TODO: provide more context of the mortality-PM10 study{]}.
Analysts may expect a PM10 coefficient between {[}0, 0.005{]} after
considering the temperature confounding {[}TODO: reference{]}. This
expectation can be framed into a check that fails, labelled as 1, if the
PM10 coefficient is outside the range {[}0, 0.005{]}. Multiple factors
can impact the PM10 coefficient, such as the sample size, the strength
of the correlation between mortality and PM10, and the strength of the
correlation between mortality and temperature. Analysts may expect a
reasonable sample size to ensure the reliability of the coefficient
estimate. Outliers in the three variables can also leverage the
coefficient. While these are possible factors that could affect the
analysis result, it is not clear the cutoff values for these checks to
determine a failure. Here a list of checks are created in
Table~\ref{tbl-checks}.

\begin{longtable}[]{@{}l@{}}
\caption{A list of checks considered for the generalized linear model of
mortality on PM10 and temperature. The checks are based on the sample
size, correlation between mortality and PM10, correlation between
mortality and temperature, and univariate outlier detection. Multiple
cutoff values are specified for each check to determine a
failure.}\label{tbl-checks}\tabularnewline
\toprule\noalign{}
the check fails if \ldots{} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
the check fails if \ldots{} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Sample size less than or equal to 200 \\
Sample size less than or equal to 400 \\
Sample size less than or equal to 600 \\
Sample size less than or equal to 800 \\
Mortality-PM10 correlation less than -0.03 \\
Mortality-PM10 correlation less than -0.04 \\
Mortality-PM10 correlation less than -0.05 \\
Mortality-PM10 correlation less than -0.06 \\
Mortality-temperature correlation greater than -0.3 \\
Mortality-temperature correlation greater than -0.35 \\
Mortality-temperature correlation greater than -0.4 \\
Mortality-temperature correlation greater than -0.45 \\
Outlier(s) are presented in the variable PM10 \\
Outlier(s) are presented in the variable mortality \\
\end{longtable}

To generate replicate of the data, we first simulate the correlation
matrix of the three variables in a grid and then use a Gaussian copula
to generate a multivariate normal distribution based on the specified
correlation matrix and sample size. The multivariate normal distribution
is transformed using the normal CDF before the inverse CDF of the
assumed distributions of the three variables is applied. To determine
the appropriate distribution of each variable, various distributions are
fitted and compared. This includes poisson and negative binomial for
mortality; gamma, log-normal, exponential, weibull, and normal for pm10
and temperature; and beta for pm10 after rescaling the data to
\([0-1]\). AIC is used to determine the best distribution fit for each
variable with qq plot presented in Figure~\ref{fig-dist-fit} to evaluate
the fit. AIC suggests a negative binomial distributio nfor mortality, a
beta distribution for PM10 (multiple by 100 to obtain the original
scale), and a Weibull distribution for temperature. To include the
effect of outlier, we add a single outlier to the data for mortality and
PM10 {[}more details{]}.

@@

\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedleft
tests
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
precision
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
recall
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
overlapping
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
independence
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
harmonic
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
arithmetic
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
1 & 0.087 & 0.215 & 0 & 1 & 0.175 & 0.434 \\
2 & 0.988 & 0.610 & 0 & 1 & 0.822 & 0.866 \\
3 & 0.392 & 0.581 & 0 & 1 & 0.569 & 0.658 \\
4 & 0.649 & 0.641 & 0 & 1 & 0.732 & 0.763 \\
5 & 0.760 & 0.880 & 0 & 1 & 0.869 & 0.880 \\

@@

A logic regression is fitted to predict whether the PM10 coefficient is
unexpected (outside the range of {[}0, 0.005{]}) using the checks listed
in Table~\ref{tbl-checks}. Precision, recall, and independence score,
along with their harmonic and arithmetic mean are calculated.
Figure~\ref{fig-linear-reg-tree} shows the logic regression tree from
the fitted model and Table~\ref{tbl-linear-reg} prints the numerical
summary of four selected single test and their combined test found by
the logic regression. The logic regression model picks up the following
cutoff value for each type of check:

\begin{itemize}
\tightlist
\item
  sample size \emph{larger than} 200
\item
  mortality-temperature correlation greater than -0.35
\item
  mortality-PM10 correlation less than -0.04
\item
  mortality contain outliers that are detected by the univariate outlier
  detection
\end{itemize}

The tree structure suggests checking mortality-PM10 correlation and a
sample size larger than 200 with an additional check of either outlier
on mortality or correlation between mortality and temperature. This
combined check rule generates a 0.76 precision and a 0.88 recall for
predicting the unexpected PM10 coefficient. The single check,
\texttt{cor(m,\ PM10)\ \textless{}\ -0.03}, is also powerful with a high
precision of 0.988, but the low recall value of 0.61 suggests its high
false positive rate, as compared to the combined rule suggested by the
logic regression.

### 2024-12-05 index.qmd
update the linear regression example

The rest of the paper is organized as follows: @sec-lit-review reviews the concept of diagnosing unexpected outcomes and data quality checks. @sec-plan introduces the concept of analysis plan, illustrated with a toy example on fitness step count. @sec-method describes the procedure for selecting optimal check combinations to predict unexpected outcomes. @sec-examples applies this selection procedure to a linear regression example examining the effect of PM10 on mortality. @sec-discussion discusses a few key considerations and @sec-conclusion concludes the paper.

@@

A regression example is produced to illustrate the test selection process for analyzing the effect of PM10 on mortality. The example demonstrates how the test process can be used to select cutoff value in the unit tests and how the test can be used to diagnose the unexpected results.

## Linear regression {#sec-linear-reg}

Consider a linear regression model to study the effect of PM10 on mortality (provide context of using PM10 to study mortality). From the literature, analysts may expect a significant (p-value $\le$ 0.05) PM10 coefficient in the linear model after considering the temperature confounding \[TODO add reference\]. This expectation can be framed into a unit test: the test fails, labelled as 1, if the p-value of the PM10 coefficient is greater than 0.05. And we have devised three unit tests to consider the 1) sample size, 2) strength of correlation between mortality and PM10, and 3) strength of correlation between mortality and temperature. \[sample size matters because ..., correlation between mortality and PM10 matters because...\]. However, it is not clear how to set the cutoff value for these tests to determine a failure. Here a list of tests are created in @tbl-tests.

| the test fails if ... | 
|--------------| 
| sample size less than or equal to 200 |
| sample size less than or equal to 400 |
| sample size less than or equal to 600 |
| sample size less than or equal to 800 |
|  mortality-PM10 correlation less than -0.03 |
|  mortality-PM10 correlation less than -0.04 |
|  mortality-PM10 correlation less than -0.05 |
|  mortality-PM10 correlation less than -0.06 |
| mortality-temperature correlation greater than -0.3 | 
| mortality-temperature correlation greater than -0.35 | 
| mortality-temperature correlation greater than -0.4 | 
| mortality-temperature correlation greater than -0.45 |

: A list of unit tests created for the linear regression model of mortality on PM10 and temperature. The tests are based on the sample size, correlation between mortality and PM10, and correlation between mortality and temperature. The cutoff values for each test are specified to determine a failure. {#tbl-tests}

To generate replicate of the data, we first simulate the correlation matrix for the three variables: mortality, PM10, and temperature with the following vector:

-   mortality and PM10: -0.03 to -0.1
-   mortality and temperature: -0.2 to -0.4
-   PM10 and temperature: 0.2 to 0.4

We then use a Gaussian copula to generate a multivariate normal distribution based on the specified correlation matrix and sample size. The multivariate normal distribution is transformed using the normal CDF, followed by the inverse CDF of the assumed distributions of the three variables. Various distributions are fitted to the variables, including poisson and negative binomial for mortality; gamma, log-normal, exponential, weibull, and normal for pm10 and temperature; and beta for pm10 after rescaling the data to $[0-1]$. AIC is used to determine the best distribution fit for each variable with qq plot presented in @fig-dist-fit to evaluate the fit. The following distributions and parameters, estimated from the original data, are used to simulate the three variables:

-   mortality, negative binomial distribution with size = 74 and mu = 183
-   PM10, beta distribution with shape1 = 4.21 and shape2 = 11.67, then multiply by 100
-   temperature, weibull distribution with shape = 3.8 and scale = 61

@@

```{r}
# cor(pm10$pm10, pm10$mortality)
# cor(pm10$temp, pm10$mortality)
# cor(pm10$temp, pm10$pm10)
# Generate correlation matrices 
corr_grid <- expand.grid(seq(0.01, 0.15, by = 0.02),                         
                         seq(0.25, 0.5, by = 0.05),                           
                         seq(0.25, 0.45, by = 0.05))

# Function to compute correlation matrix for each combination 
gen_corr_mtx <- function(r1, r2, r3) {   
  # correlation between mortality and pm10 are negative - r1
  # correlation between mortality and temp are negative - r2
  cor_matrix <- matrix(c(1, -r1, -r2,                          
                         -r1, 1, r3,                          
                         -r2, r3, 1), nrow = 3, byrow = TRUE)          
  # Store the matrix in the list   
  if (all(eigen(cor_matrix)$values > 0)) return(cor_matrix) 
  }

# Plan for parallel processing
corr_mtx <- lapply(1:nrow(corr_grid), function(i) {   
  gen_corr_mtx(corr_grid[i, 1], corr_grid[i, 2], corr_grid[i, 3]) }) 
corr_mtx <- corr_mtx[map_lgl(corr_mtx, ~!is.null(.x))]  
sample_size <- c(400, 600, 800) 
model <- c("mortality ~ pm10 + temp") |> map(as.formula)

generate_data <- function(n, mtx, seed = 123, outlier = FALSE) {   
  mu <- c(0, 0, 0)   
  set.seed(seed)  
  data <- mvrnorm(n, mu, mtx, empirical = TRUE)   
  U <- pnorm(data, mean = 0, sd = 1)   
  tibble(mortality = qnbinom(U[,1], size = 74, mu = 183),           
         pm10 = qbeta(U[,2], shape1 = 4.21, shape2 = 11.67) * 100,           
         temp = qweibull(U[,3], shape = 3.8, scale = 61))   
}

res <- tibble(corr_mtx = corr_mtx) |> 
  mutate(id = row_number()) |> 
  crossing(sample_size, model) |> 
  rowwise() |>   
  mutate(data = list(generate_data(n = sample_size, mtx = corr_mtx)),          
         fit = list(summary(lm(model, data))$coefficients))

pm10_df <- res |>   
  mutate(     
    p_value = fit[2,4],     
    coef = fit[2,1],     
    mor_pm10_correlation = corr_mtx[1, 2],     
    mor_temp_correlation = corr_mtx[1, 3],     
    temp_pm10_yz_correlation = corr_mtx[2, 3],     
    fml = deparse(model) |> as.factor(),     
    unexpect = ifelse(p_value < 0.05, 0, 1),
    #unexpect = ifelse(between(coef, 0, 0.15), 0, 1),
    `smpl_size <= 200` = ifelse(sample_size <= 200, 1, 0),
    `smpl_size <= 400` = ifelse(sample_size <= 400, 1, 0),
    `smpl_size <= 600` = ifelse(sample_size <= 600, 1, 0),
    `smpl_size <= 800` = ifelse(sample_size <= 800, 1, 0),
    `cor(m, PM10) < -0.03` = ifelse(mor_pm10_correlation < -0.03, 1, 0),
    `cor(m, PM10) < -0.04` = ifelse(mor_pm10_correlation < -0.04, 1, 0),
    `cor(m, PM10) < -0.05` = ifelse(mor_pm10_correlation < -0.05, 1, 0),
    `cor(m, PM10) < -0.06` = ifelse(mor_pm10_correlation < -0.06, 1, 0),
    `cor(m, tmp) > -0.3` = ifelse(mor_temp_correlation > -0.3, 1, 0),
    `cor(m, tmp) > -0.35` = ifelse(mor_temp_correlation > -0.35, 1, 0),
    `cor(m, tmp) > -0.4` = ifelse(mor_temp_correlation > -0.4, 1, 0),
    `cor(m, tmp) > -0.45` = ifelse(mor_temp_correlation > -0.45, 1, 0)
  ) |>    
  ungroup()

@@

```{r}
#| fig.width: 8
#| fig.heigh: 3
fit <- pm10_df |> fit_logic_reg(unexpect, `smpl_size <= 200`:`cor(m, tmp) > -0.45`, 
                                seed = 4, nleaves = 3)
plot_logicreg(fit) 
```
<!-- The caption with some LaTeX like this $\lambda$. -->

```{r}
#| label: tbl-linear-reg
library(kableExtra)
df <- list(
  tibble(.fitted = pm10_df$`smpl_size <= 600`, 
         unexpect = pm10_df$unexpect),
  tibble(.fitted = pm10_df$`cor(m, PM10) < -0.04`, 
         unexpect = pm10_df$unexpect),
  tibble(.fitted = pm10_df$`cor(m, tmp) > -0.45`, 
         unexpect = pm10_df$unexpect),
  augment(fit)) |>
  map_dfr(~.x |> calc_miscla_rate(unexpect, .fitted) |> 
            mutate(overlapping = 0, independence = 1) |> 
            calc_metrics(metrics = c("harmonic", "arithmetic")), .id = "id")

tbl <- tibble(tests = c(1,2,3,4)) |> bind_cols(df |> dplyr::select(-id)) 
tbl |> 
  kable(digits = 3, escape = FALSE,
        caption = "Accuracy (precision and recall) and parsimony (independence) metrics derived from the logic regression model, along with harmonic and arithmetic means, for individual unit tests (1: sample size, 2: mortality-PM10 correlation, 3: mortality-temperature correlation), and the combined test rule 4: (sample size AND mortality-temperature correlation) OR mortality-PM10 correlation.")

A total of `r ncol(pm10_df) - ncol(res) - 7` tests with four cutoff values each for the three types of tests are used to fit the logic regression and calculate the precision and recall score for predictability of an unexpected outcome (p value \< 0.05). @fig-linear-reg-tree shows the logic regression tree from the fitted model and @tbl-linear-reg prints the numerical summary of three selected single test and their combined test found by the logic regression. The logic regression model is capable of picking up the best cutoff value for each type of tests and yields the following three tests:

-   sample size less than or equal to 600
-   mortality-temperature correlation greater than -0.35
-   mortality-PM10 correlation less than -0.05

The combined test rule, (mortality-tempearture correlation AND sample size) OR mortality-PM10 correlation, has the highest precision, recall, and independence, suggesting that it is the most effective test for diagnosing unexpected results in the linear regression model given our assumption on the data. The tree structure suggests obtaining the mortality-PM10 correlation is as important as obtaining a combined relationship between mortality-temperature correlation and sample size in diagnosing unexpected results in the linear regression model. And based on our assumption of the variable distribution from the data simulation, we can obtain a precision of 0.908 and a recall of 0.794 for the unexpected outcome. These provides check points for the analysts to diagnose the unexpected results and to improve the model.

>>>>

The rest of the paper is organized as follows: @sec-lit-review reviews the concept of diagnosing unexpected outcomes and data quality checks. @sec-plan introduces the concept of analysis plan, illustrated with a toy example on fitness step count. @sec-method describes the procedure for selecting optimal check combinations to predict unexpected outcomes. @sec-pm10-mortality applies this selection procedure to a linear regression example examining the effect of PM10 on mortality. @sec-discussion discusses a few key considerations and @sec-conclusion concludes the paper.

@@

A regression example is produced to illustrate the test selection process for analyzing the effect of PM10 on mortality. The example demonstrates how the test process can be used to select cutoff values in the unit tests and how the test can be used to diagnose an unexpected PM10 coefficient from the generalized linear model.

## Effect of PM10 on mortality {#sec-pm10-mortality}

Consider a generalized linear model (GLM) to study the effect of PM10 on mortality [TODO: provide more context of the mortality-PM10 study]. Analysts may expect a PM10 coefficient between [0, 0.005] after considering the temperature confounding [TODO: reference]. This expectation can be framed into a check that fails, labelled as 1, if the PM10 coefficient is outside the range [0, 0.005]. Multiple factors can impact the PM10 coefficient, such as the sample size, the strength of the correlation between mortality and PM10, and the strength of the correlation between mortality and temperature. Analysts may expect a reasonable sample size to ensure the reliability of the coefficient estimate. Outliers in the three variables can also leverage the coefficient. While these are possible factors that could affect the analysis result, it is not clear the cutoff values for these checks to determine a failure. Here a list of checks are created in @tbl-checks.

| the check fails if ... | 
|--------------| 
| Sample size less than or equal to 200 |
| Sample size less than or equal to 400 |
| Sample size less than or equal to 600 |
| Sample size less than or equal to 800 |
| Mortality-PM10 correlation less than -0.03 |
| Mortality-PM10 correlation less than -0.04 |
| Mortality-PM10 correlation less than -0.05 |
| Mortality-PM10 correlation less than -0.06 |
| Mortality-temperature correlation greater than -0.3 | 
| Mortality-temperature correlation greater than -0.35 | 
| Mortality-temperature correlation greater than -0.4 | 
| Mortality-temperature correlation greater than -0.45 | 
| Outlier(s) are presented in the variable PM10 | 
| Outlier(s) are presented in the variable mortality |

: A list of checks considered for the generalized linear model of mortality on PM10 and temperature. The checks are based on the sample size, correlation between mortality and PM10, correlation between mortality and temperature, and univariate outlier detection. Multiple cutoff values are specified for each check to determine a failure. {#tbl-checks}

To generate replicate of the data, we first simulate the correlation matrix of the three variables in a grid and then use a Gaussian copula to generate a multivariate normal distribution based on the specified correlation matrix and sample size. The multivariate normal distribution is transformed using the normal CDF before the inverse CDF of the assumed distributions of the three variables is applied. To determine the appropriate distribution of each variable, various distributions are fitted and compared. This includes poisson and negative binomial for mortality; gamma, log-normal, exponential, weibull, and normal for pm10 and temperature; and beta for pm10 after rescaling the data to $[0-1]$. AIC is used to determine the best distribution fit for each variable with qq plot presented in @fig-dist-fit to evaluate the fit. AIC suggests a negative binomial distributio nfor mortality, a beta distribution for PM10 (multiple by 100 to obtain the original scale), and a Weibull distribution for temperature. To include the effect of outlier, we add a single outlier to the data for mortality and PM10 [more details].

@@

```{r}
# cor(pm10$pm10, pm10$mortality)
# cor(pm10$temp, pm10$mortality)
# cor(pm10$temp, pm10$pm10)
corr_grid <- expand.grid(seq(0.01, 0.2, by = 0.02),                         
                         seq(0.25, 0.5, by = 0.05),                           
                         seq(0.25, 0.45, by = 0.05))

gen_corr_mtx <- function(r1, r2, r3) {   
  # correlation between mortality and pm10 are negative - r1
  # correlation between mortality and temp are negative - r2
  cor_matrix <- matrix(c(1, -r1, -r2,                          
                         -r1, 1, r3,                          
                         -r2, r3, 1), nrow = 3, byrow = TRUE)          
  if (all(eigen(cor_matrix)$values > 0)) return(cor_matrix) 
  }

corr_mtx <- lapply(1:nrow(corr_grid), function(i) {   
  gen_corr_mtx(corr_grid[i, 1], corr_grid[i, 2], corr_grid[i, 3]) }) 
corr_mtx <- corr_mtx[map_lgl(corr_mtx, ~!is.null(.x))]  
sample_size <- c(200, 400, 600, 800, 1000)
outlier <- c(TRUE, FALSE)

generate_data <- function(n, mtx, seed = 123, outlier = FALSE) {   
  mu <- c(0, 0, 0)   
  set.seed(seed)  
  data <- mvrnorm(n, mu, mtx, empirical = TRUE)   
  U <- pnorm(data, mean = 0, sd = 1)

  if (!outlier) {
   tibble(mortality = qnbinom(U[,1], size = 74, mu = 183),           
         pm10 = qbeta(U[,2], shape1 = 4.21, shape2 = 11.67) * 100,           
         temp = qweibull(U[,3], shape = 3.8, scale = 61))    
  } else{
    pm10_vec <- qbeta(U[,2], shape1 = 4.21, shape2 = 11.67) * 100
    tibble(
      mortality = c(qnbinom(U[,1], size = 74, mu = 183)[-1],
                    rnorm(n = 1, mean = 300, sd = 10)),           
      pm10 = c(pm10_vec[-1], rnorm(n = 1, mean = 100, sd = 10)),           
      temp = qweibull(U[,3], shape = 3.8, scale = 61)
      )    
  }

}

res <- tibble(corr_mtx = corr_mtx) |> 
  mutate(id = row_number()) |> 
  crossing(sample_size, outlier) |> 
  rowwise() |>   
  mutate(data = list(generate_data(n = sample_size, mtx = corr_mtx, outlier = outlier)),          
         fit = list(summary(glm(mortality ~ pm10 + temp, family = "poisson", data))$coefficients))

pm10_df <- res |>   
  mutate(     
    p_value = fit[2,4],     
    coef = fit[2,1],     
    mor_pm10_correlation = corr_mtx[1, 2],     
    mor_temp_correlation = corr_mtx[1, 3],     
    temp_pm10_yz_correlation = corr_mtx[2, 3],     
    unexpect = ifelse(between(coef, 0, 0.005), 0, 1),
    `smpl_size <= 200` = ifelse(sample_size <= 200, 1, 0),
    `smpl_size <= 400` = ifelse(sample_size <= 400, 1, 0),
    `smpl_size <= 600` = ifelse(sample_size <= 600, 1, 0),
    `smpl_size <= 800` = ifelse(sample_size <= 800, 1, 0),
    `cor(m, PM10) < -0.03` = ifelse(mor_pm10_correlation < -0.03, 1, 0),
    `cor(m, PM10) < -0.04` = ifelse(mor_pm10_correlation < -0.04, 1, 0),
    `cor(m, PM10) < -0.05` = ifelse(mor_pm10_correlation < -0.05, 1, 0),
    `cor(m, PM10) < -0.06` = ifelse(mor_pm10_correlation < -0.06, 1, 0),
    `cor(m, tmp) > -0.3` = ifelse(mor_temp_correlation > -0.3, 1, 0),
    `cor(m, tmp) > -0.35` = ifelse(mor_temp_correlation > -0.35, 1, 0),
    `cor(m, tmp) > -0.4` = ifelse(mor_temp_correlation > -0.4, 1, 0),
    `cor(m, tmp) > -0.45` = ifelse(mor_temp_correlation > -0.45, 1, 0)
  ) |>    
  ungroup()

pm10_df <- pm10_df |> 
  rowwise() |> 
  mutate(`PM10 outlier` = ifelse(any(scale(data$pm10) > 4), 1, 0),
         `mortality_outlier` = ifelse(any(scale(data$mortality) > 4), 1, 0)
         ) |> 
  ungroup()
```

```{r}
#| fig.width: 8
#| fig.heigh: 3
fit <- pm10_df |> fit_logic_reg(unexpect, `smpl_size <= 200`:`mortality_outlier`, 
                                seed = 6, nleaves = 4)
plot_logicreg(fit) 
```
<!-- The caption with some LaTeX like this $\lambda$. -->

```{r}
#| label: tbl-linear-reg
library(kableExtra)
df <- list(
  tibble(.fitted = pm10_df$`smpl_size <= 200`, 
         unexpect = pm10_df$unexpect),
  tibble(.fitted = pm10_df$`cor(m, PM10) < -0.03`, 
         unexpect = pm10_df$unexpect),
  tibble(.fitted = pm10_df$`cor(m, tmp) > -0.35`, 
         unexpect = pm10_df$unexpect),
  tibble(.fitted = pm10_df$mortality_outlier, 
         unexpect = pm10_df$unexpect),
  augment(fit)) |>
  map_dfr(~.x |> calc_miscla_rate(unexpect, .fitted) |> 
            mutate(overlapping = 0, independence = 1) |> 
            calc_metrics(metrics = c("harmonic", "arithmetic")), .id = "id")

tbl <- tibble(tests = c(1,2,3,4, 5)) |> bind_cols(df |> dplyr::select(-id)) 
tbl |> 
  kable(digits = 3, escape = FALSE,
        caption = "Accuracy (precision and recall) and parsimony (independence) metrics derived from the logic regression model, along with harmonic and arithmetic means, for individual unit tests (1: sample size, 2: mortality-PM10 correlation, 3: mortality-temperature correlation), and the combined test rule 4: (sample size AND mortality-temperature correlation) OR mortality-PM10 correlation.")

@@

A logic regression is fitted to predict whether the PM10 coefficient is unexpected (outside the range of [0, 0.005]) using the checks listed in @tbl-checks. Precision, recall, and independence score, along with their harmonic and arithmetic mean are calculated. @fig-linear-reg-tree shows the logic regression tree from the fitted model and @tbl-linear-reg prints the numerical summary of four selected single test and their combined test found by the logic regression. The logic regression model picks up the following cutoff value for each type of check:

-   sample size *larger than* 200
-   mortality-temperature correlation greater than -0.35
-   mortality-PM10 correlation less than -0.04
-   mortality contain outliers that are detected by the univariate outlier detection

<!-- (mortality-temperature correlation > -0.35 OR there exist mortality outlier) and (mortality-PM10 correlation < -0.04 AND (not sample size less than 200) )  -->

The tree structure suggests checking mortality-PM10 correlation and a sample size larger than 200 with an additional check of either outlier on mortality or correlation between mortality and temperature. This combined check rule generates a `r round(as.numeric(df[5, "precision"]), 3)` precision and a `r round(as.numeric(df[5, "recall"]), 3)` recall for predicting the unexpected PM10 coefficient. The single check, `cor(m, PM10) < -0.03`, is also powerful with a high precision of `r round(as.numeric(df[2, "precision"]), 3)`, but the low recall value of `r round(as.numeric(df[2, "recall"]), 3)` suggests its high false positive rate, as compared to the combined rule suggested by the logic regression.
